---
title: "2024 NPT Water Master Graphs"
output: html_document
date: "2025-01-08"
author: "Maille Gray"
---
#Libraries
```{r}
library(tidyverse) #includes ggplot, dplyr, lubridate, purrr
library(openxlsx)
```

#Read the .csv. If unsure of your working directory, use the getwd() and setwd() functions.
```{r}
#read the .csv
df <- read.csv("NPTsummary.csv") #make sure the .csv is in the same working directory

#convert the Date column from a string to a date
df$Date <- as.Date(df$Date)

```

#Identify acequia
```{r}
#Name of the desired acequia goes here
acequia = "CICD"
```

#calculate statistics for the acequia
```{r}
#replace space with . in acequia name to make compatible with dataframe column names
colname <- str_replace_all(acequia, " ", ".")

#dataframe "a" holds the date, the daily volume, and accumulation of daily volume
a <- data.frame(
  date = as.Date(df$Date), #read Date column as a date type
  daily = df[[colname]],  # Using [[ to dynamically reference the column
  total = cumsum(df[[colname]])  # Use cumsum to accumulate the values
)

#make a table of the monthly totals
summary <- aggregate(cbind(daily)~month(date),
                     data=a,FUN=sum)
colnames(summary) <- c("month", "total") #rename columns
summary <- summary %>% mutate(month = month.abb[as.numeric(month)]) #replace the month number with month abbrev.

#calculate the total diversion and create a summary table
full <- a %>%
  summarise(full = sum(daily, na.rm = TRUE))
pull (full) #save the total as numeric value
summary[nrow(summary) + 1,] = c("Total",full) #add the total to the summary column

#switch rows and columns and save in s
s <- t(summary)

#write table to excel
filename = paste(acequia, ".xlsx")
write.xlsx(s,filename, rowNames=FALSE, colNames=FALSE)
```

#Make the diversion graph. Because we're working with two different scales, one for the daily diversion and one for the total diversion, there's a scaling coefficiant so that the lines plot evenly. The Total Volume amount is multiplied by this coefficiant so it plots at the same scale as the Daily Diversion, then the legend amounts are divided by the coefficiant so they reflect the actual values.
```{r}
#scaling coefficiant
coefficiant = 0.01

#make the ggplot using dataframe a
plot <- ggplot(a) + 
  #plot the daily diversion bars. color is set to what we want to appear in the legend. items within the aes() brackets appear in the legend, but items outside won't show up in the legend
  geom_col(aes(x = date, y = daily, colour = "Daily Diversion", width = 0.05), fill = "#47ACF1") +
  #plot total diversion line. Note multiplying coefficient to make scales more even
  geom_line(aes(x = date, y = coefficiant*total, colour = "Total Diversion"), linewidth = 1.2, group = 1) +
  #use scale_color_manual to assign a color to the legend items
  scale_color_manual(values = c("Daily Diversion"="#4594ca", "Total Diversion"="darkblue"))+
  scale_y_continuous(sec.axis = sec_axis(~./coefficiant, name = "Total Volume"))+ #set up second axis
  #set up x axis labels. The date breaks occur every one month. %b calls the month abbreviation and %d calls the day
  scale_x_date(date_breaks = "1 month", date_labels= "%b %d")+
  labs(y = "Daily Volume", title = paste(acequia, "Daily Average & Total Volume (Acre-Feet)"))+ #set up y labels and title
  theme_bw()+ #black/white theme
  theme(
    text = element_text(family = "serif"), #set font to Times
    legend.position = "bottom", #put legend at bottom
    legend.title = element_blank(), #remove legend title
    axis.title.x=element_blank(), #remove x axis title
    plot.title = element_text(hjust = 0.5, size = 14)) #move and resize plot title
plot
```

#save plot
```{r}
#save the plot as a .jpg
ggsave(plot,
       filename = paste(acequia, ".jpg"),
       device = "jpg",
       height = 3.25, width = 5, units = "in" )
```

##Use in unique cases and overdiversions

#Nambe manipulations
#not currently using
```{r}
#create a dataframe only using Highline, Upper Consolidated, and Lower Consolidated
n <- data.frame(
  date = as.Date(df$Date), #read Date column as a date type
  Highline = df$Highline,
  Upper.Consolidated = df$Upper.Consolidated,
  Lower.Consolidated = df$Lower.Consolidated,
  h = cumsum(df$Highline),
  u = cumsum(df$Upper.Consolidated),
  l = cumsum(df$Lower.Consolidated) # Use cumsum to accumulate the values
)
n$sum <- n$h + n$u + n$l #make a sum column of all the diversions

#set threshold to allowed Pueblo diversion
threshold <- 1459
#Find the first instance where diversion is met
result <- n %>%
  filter(sum > threshold) %>%
  slice(1) %>%  # Select the first row where the max diversion is met
  select(date, h, u, l)

#create a column that indicates when overdiversion has been reached
n <- n %>%
  mutate(h_exceeds_threshold = ifelse(h > result$h, "Exceeds Max Allowable", " Total Volume"))


coefficiant = 0.01 #adjust scaling of bars and line

#make the plot for Highline
highline <- ggplot(n) + 
  geom_col(aes(x = date, y = Highline, colour = "Daily Diversion", width = 0.1), fill = "#47ACF1") + #plot the daily diversion as columns
  scale_color_manual(
    values = c("Exceeds Max Allowable" = "red", 
               " Total Volume" = "darkblue", 
               "Daily Diversion" = "#4594ca"),
    breaks = c("Daily Diversion", "Exceeds Max Allowable", " Total Volume")  # Specify the order
  ) + 
  #plot total diversion line. Note multiplying coefficient to make scales more even. Color is based on the "exceeds threshold" column
  geom_line(aes(x = date, y = coefficiant * h, colour = h_exceeds_threshold), linewidth = 1.2, group = 1) + # Color line based on h_exceeds_threshold
  scale_y_continuous(sec.axis = sec_axis(~./coefficiant, name = "Total Volume"))+ #set up second axis
  scale_x_date(date_breaks = "1 month", date_labels= "%b %d")+ #set up x axis labels
  labs(y = "Daily Volume", title = "Highline Daily Average & Total Volume (Acre-Feet)")+ #set up y labels
  theme_bw()+ #black/white theme
  geom_segment(data = result, aes(x = date, y = coefficiant * h, xend = date+3, yend = coefficiant * h-0.5, colour = "segment"))+
  geom_point(data = result, aes(x = date, y = coefficiant * h), color = "red", size = 4) + 
  geom_label(data = result, aes(x = date+3, y = coefficiant * h-0.5, label = format(date, "%b %d")), 
            vjust = 1, hjust = 0, size = 3, color = "black")+
  theme(
    text = element_text(family = "serif"), #set font to Times
    legend.position = "bottom", #put legend at bottom
    legend.title = element_blank(), #remove legend title
    axis.title.x=element_blank(), #remove x axis title
    plot.title = element_text(hjust = 0.5, size = 18)) #move and resize plot title
highline
```

#plot graphs for acequias that overdiverted
```{r}
threshold = 51.43 #enter allowed diversion amount

#create a column called exceeds_threshold to identify points where allowed diversion is exceeded
a <- a %>%
  mutate(exceeds_threshold = ifelse(a$total > threshold, "Exceeds Max Allowable", " Total Volume"))

#find the first date where maximum allowed diversion is exceeded
result <- a %>%
  filter(total > threshold) %>%
  slice(1) %>%  # Select the first row where the condition is met
  select(date, total)

coefficiant = 0.012 #adjusts scaling of bars and line

#make the plot so that the total volume line turns red when max diversion is reached
p <- ggplot(a) + 
  geom_col(aes(x = date, y = daily, colour = "Daily Diversion", width = 0.1), fill = "#47ACF1") + #plot the daily diversion as columns
  #plot total diversion line. Note multiplying coefficient to make scales more even. Line color is based on whether or not the diversion exceeds the maximum
  geom_line(aes(x = date, y = coefficiant * total, colour = exceeds_threshold), linewidth = 1.2, group = 1) +
  #set legend colors
  scale_color_manual(
    values = c("Exceeds Max Allowable" = "red", 
               " Total Volume" = "darkblue", 
               "Daily Diversion" = "#4594ca"),
    breaks = c("Daily Diversion", " Total Volume", "Exceeds Max Allowable")  # Specify the legend order
  ) + 
  scale_y_continuous(sec.axis = sec_axis(~./coefficiant, name = "Total Volume"))+ #set up second axis
  scale_x_date(date_breaks = "1 month", date_labels= "%b %d")+ #set up x axis labels
  labs(y = "Daily Volume", title = paste(acequia, "Daily Average & Total Volume (Acre-Feet)"))+ #set up y labels
  theme_bw()+ #black/white theme
  geom_segment(data = result, aes(x = date, y = coefficiant * total, xend = date+3, yend = coefficiant * total-0.2, colour = "segment"))+
  geom_point(data = result, aes(x = date, y = coefficiant * total), color = "red", size = 4) + 
  geom_label(data = result, aes(x = date+3, y = coefficiant * total-0.2, label = format(date, "%b %d")), 
            vjust = 1, hjust = 0, size = 3, color = "black")+
  theme(
    text = element_text(family = "serif"), #set font to Times
    legend.position = "bottom", #put legend at bottom
    legend.title = element_blank(), #remove legend title
    axis.title.x=element_blank(), #remove x axis title
    plot.title = element_text(hjust = 0.5, size = 14)) #move and resize plot title
p
```

#save plot
```{r}
#save the plot as a .jpg
ggsave(p,
       filename = paste(acequia, ".jpg"),
       device = "jpg",
       height = 3.25, width = 5, units = "in" )
```

##The following code chunks are for making plots from other data, including precipitation and snow water equivalent

#SWE
```{r}
swe <- read.csv("swe.csv") #read the .csv

swe$percentmedian <- (swe$swe/swe$medianswe)*100 #create a percent of median column

swe$date <- as.Date(swe$date) #read the date column as a Date type

swe_filtered <- swe[swe$percentmedian >= 100, ] #filter dates where swe exceeds 100% of the median
swe_filtered <- swe_filtered %>% drop_na #exclude all na values
swe_filtered = swe_filtered[-1,] #remove the first row because it was an outlier
```

#Plot SWE
```{r}
coefficiant = 0.22 #adjust scaling of bars and line. coefficiant applies to the percent median lines

#make the ggplot
plot <- ggplot(swe) + 
  #plot the percent of median line
  geom_line(aes(x = date, y = coefficiant * percentmedian, color = "Percent of Median"), linewidth = 0.5) +
  #make the percent of median line turn green where SWE exceeds 100%
  geom_line(data = swe_filtered, aes(x = date, y = coefficiant * percentmedian), color = "green", linewidth = 0.5)+
  #plot the median swe line
  geom_line(aes(x = date, y = medianswe, colour = "1991-2020 SWE Median"), linewidth = 0.8) + #plot the median swe
  #plot the 2024 swe line
  geom_line(aes(x = date, y = swe, colour = "2024 SWE"), linewidth = 1.2) +
  #plot 100% as a dotted line
  geom_line(aes(x = date, y = 100*coefficiant, color = "100% of Median"), linewidth = 0.5, linetype = "dotdash")+
  #assign colors to legend items
  scale_color_manual(
    values = c(
      "2024 SWE"="#006db7", 
      "1991-2020 SWE Median"="#41b2ff", 
      "Percent of Median" = "orange"),
    breaks = c("2024 SWE", "1991-2020 SWE Median", "Percent of Median"))+
  scale_y_continuous(sec.axis = sec_axis(~./coefficiant, name = "Percent of Median"))+ #set up second axis
  scale_x_date(date_breaks = "1 month", date_labels= "%b %d")+ #set up x axis labels
  labs(y = "Snow Water Equivalent (in)", title = paste("Snow Water Equivalent at SNOTEL Station 922 (Tesuque Peak)"))+ #set up y labels and title
  theme_bw()+ #black/white theme
  theme(
    text = element_text(family = "serif"), #set font to Times
    legend.position = "bottom", #put legend at bottom
    legend.title = element_blank(), #remove legend title
    axis.title.x=element_blank(), #remove x axis title
    plot.title = element_text(hjust = 0.5, size = 14)) #move and resize plot title
plot
```

#monthly precip values
```{r}
precip <- read.csv("precip.csv") #read the precip csv

precip$date <- as.Date(precip$date) #read date column as Date type

#manipulating the precip dataframe to make it easier to plot the snotel data as two columns
precip_long <- precip %>%
  pivot_longer(cols = c(precip_inc, precip_avg), 
               names_to = "type", 
               values_to = "precip_value") %>%
  mutate(type = factor(type, levels = c("precip_inc", "precip_avg"), 
                       labels = c("2024 Precipitation", "1991-2020 Average Precipitation")))
```

#snotel precipitation plot
```{r}
plot <- ggplot(precip_long) +
  #plot snotel precip as two bars - one for 1991-2020 average and one for 2024 monthly means
  geom_bar(aes(x = date, y = precip_value, fill = type), 
           position = "dodge", stat = "identity") +
  scale_fill_manual(
    values = c(
      "2024 Precipitation" = "#1d9bca", 
      "1991-2020 Average Precipitation" = "#7adbff"),
    breaks = c("2024 Precipitation", "1991-2020 Average Precipitation")) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + # Set up x axis labels
  labs(y = "Precipitation Increment (in)", 
       title = "Monthly Precipitation at SNOTEL Station 922 (Tesuque Peak)") + 
  theme_bw() + # Black/white theme
  theme(
    text = element_text(family = "serif"), # Set font to Times
    legend.position = "bottom", # Put legend at bottom
    legend.title = element_blank(), # Remove legend title
    axis.title.x = element_blank(), # Remove x axis title
    plot.title = element_text(hjust = 0.5, size = 14)) # Move and resize plot title

plot
```

#precip values plot
```{r}
plot <- ggplot(precip)+
  geom_line(aes(x = date, y=mean_r2,color = "Northern Mountains Area Avg"), linewidth = 1.2)+
  geom_line(aes(x=date, y =precip_inc, color = "Tesuque Peak SNOTEL Station"), linewidth = 1.2)+
  geom_line(aes(x=date, y =sfe_precip, color = "Santa Fe Airport NOAA Station"), linewidth = 1.2)+
  geom_line(aes(x = date,y=cocorahs, color = "Pojoaque CoCoRaHS Gauges"), linewidth = 1.2)+
  scale_color_manual(
    values = c(
      "Tesuque Peak SNOTEL Station" = "#5AE76D",
      "Santa Fe Airport NOAA Station" = "#EEB568",
      "Pojoaque CoCoRaHS Gauges" = "#008f98",
      "Northern Mountains Area Avg" = "#6e108c"
    )
  )+
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + # Set up x axis labels
  labs(y = "Precipitation (in)", 
       title = "2024 Monthly Precipitation") + 
  guides(color = guide_legend(nrow = 2))+
  theme_bw() + # Black/white theme
  theme(
    text = element_text(family = "serif"), # Set font to Times
    legend.position = "bottom", # Put legend at bottom
    legend.title = element_blank(), # Remove legend title
    legend.text=element_text(size=8),
    axis.title.x = element_blank(), # Remove x axis title
    plot.title = element_text(hjust = 0.5, size = 18)) # Move and resize plot title

plot
```

#save plot
```{r}
ggsave(plot,
       filename = "ncep.jpg",
       device = "jpg",
       height = 5, width = 5, units = "in" )
```

#create the ncep graph
```{r}
coefficiant = 2.5 #adjust scaling of bars and line

plot <- ggplot(precip)+
  geom_col(aes(x = date,y=anomaly_r2*coefficiant, color = "2024 Compared to 1991-2020 Mean"), fill = NA)+
  geom_line(aes(x = date, y=mean_r2,color = "2024 Area Avg"), linewidth = 1.2)+
  geom_line(aes(x=date, y =0), color = "#b688c6")+
  scale_y_continuous(sec.axis = sec_axis(~./coefficiant, name = "Deviation from Normal Conditions"))+ #set up second axis
  scale_color_manual(
    values = c(
      "2024 Compared to 1991-2020 Mean" = "#b688c6",
      "2024 Area Avg" = "#6e108c"
    )
  )+
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + # Set up x axis labels
  labs(y = "Precipitation (in)", 
       title = "NCEP/NCAR Reanalysis New Mexico : Northern Mountains \n2024 Monthly Precipitation") + 
  theme_bw() + # Black/white theme
  theme(
    text = element_text(family = "serif"), # Set font to Times
    legend.position = "bottom", # Put legend at bottom
    legend.title = element_blank(), # Remove legend title
    axis.title.x = element_blank(), # Remove x axis title
    plot.title = element_text(hjust = 0.5, size = 12)) # Move and resize plot title
plot
```